A weakly structured stem for human origins in Africa

### 2022-01-05 paper.tex
fill in discussion

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}

@@

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135ka, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral Homo spp. populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
\sgcomment{IF we were not afraid of the racists, we could write something like:
"Despite the genetic similarity between these populations, 4 to 10\% of random
genetic differences among [modern human populations] can be attributed to this
early genetic drift.}
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar. We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
distributed across Africa, and are particularly well represented in the
northern, eastern and southern parts of the continent. Similarly, fossil crania
such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
anatomically derived \emph{Homo sapiens} features were also present across the
continent during this period. It has been difficult to reconcile these lines of
evidence with evidence from genomics, which have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population. It is unclear whether fossil specimens and archaeological sites
represent populations which contributed to our ancestors as population
precedents, or were local “dead-ends” from which contemporary \emph{Homo
sapiens} do not descend. Recently, synthetic attempts to reconcile genetic and
paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of Homo sapiens beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (MSL from Sierre Leone
\citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz (recent descendents of a
hunter-gatherer group from Ethiopia \citep{Gurdasani2015-qy,Gopalan2019-wd}),
and Eastern African agriculturalists (Amhara and Oromo from Ethiopia
\citep{Gurdasani2015-qy}). The Amhara and Oromo populations, despite speaking
distinct Afro-Asiatic languages, are highly genetically similar
\citep{Pagani2015-pz,Gopalan2019-wd} and thus the two groups were combined for
a larger sample size (Figure~\ref{fig:1}). We also included the British
(GBR) from the 1000 Genomes Project in our demographic models as a
representative source of back-to-Africa gene flow and recent colonial admixture
in South Africa. Finally, we added a high-coverage ancient Neanderthal genome
from Vindija Cave, Croatia \citep{Prufer2017-kk} to account for archaic gene
flow from Neanderthals in non-Africans and gauge the relative time depth of
divergence, assuming Neanderthals diverged 550ka from a common stem. For each
population, we computed low-order allele frequency and linkage disequilibrium
(LD) statistics that are well suited for both low- and high-coverage genomes
\citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a maximum-likelihood inference
framework, we then fit to these statistics a family of parameterized
demographic models that involve population splits, size changes, continuous and
variable migration rates, and punctuated admixture events, to learn about the
nature of population structure over the past million years.

@@

Regardless of the model choice for early epochs, inference of recent human
demographic history from two thousand \aprcomment{why two thousand, instead of
just ``up to 150ka''?} to 150ka was remarkably robust. The earliest divergence
among contemporary sampled human populations differentiates the southern
African Nama from other African groups between 110--135ka, with low to moderate
levels of subsequent gene flow \aprcomment{(Table 1)}. In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$150ka. We conclude that, regardless of model
specification, geographic patterns of \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems \aprcomment{(Figure 3)}.

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (Supp.
Methods). The time of divergence of Western and Eastern African populations was
set to 60ka, just prior to the split of Eurasians and East Africans 50ka. Both
models predict relatively high subsequent gene flow between the two regions
($m=2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by ADMIXTURE
\aprcomment{cite} (Figure~\ref{fig:1}).

@@

@@

\section*{Methods}
A placeholder citation \citep{Kelleher2016-lw}.

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{} % \includegraphics{figures/fig2.png}}
\caption{\textbf{The second main figure}.
    A placeholder - best fit model(s).
}
\label{fig:2}
\end{center}
\end{figure}

@@

@@

>>>>

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}

@@

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral Homo spp. populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
\sgcomment{IF we were not afraid of the racists, we could write something like:
"Despite the genetic similarity between these populations, 4 to 10\% of random
genetic differences among [modern human populations] can be attributed to this
early genetic drift.}
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar. We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

Archaeological sites from the Middle Stone Age (approx. 300kya-40kya) are widely
distributed across Africa, and are particularly well represented in the
northern, eastern and southern parts of the continent. Similarly, fossil crania
such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
anatomically derived \emph{Homo sapiens} features were also present across the
continent during this period. It has been difficult to reconcile these lines of
evidence with evidence from genomics, which have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population. It is unclear whether fossil specimens and archaeological sites
represent populations which contributed to our ancestors as population
precedents, or were local “dead-ends” from which contemporary \emph{Homo
sapiens} do not descend. Recently, synthetic attempts to reconcile genetic and
paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of Homo sapiens beginning at least 300kya
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (MSL from Sierre Leone, Phase 3 of
the 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz
(recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we added a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for archaic gene flow from Neanderthals in non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550kya from a common
stem. For each population, we computed low-order allele frequency and linkage
disequilibrium (LD) statistics that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

Regardless of the model choice for early epochs, inference of recent human
demographic history from two thousand \aprcomment{why two thousand, instead of
just ``up to 150kya''?} to 150kya was remarkably robust. The earliest divergence
among contemporary sampled human populations differentiates the southern
African Nama from other African groups between 110--135kya, with low to moderate
levels of subsequent gene flow \aprcomment{(Table 1)}. In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$150kya. We conclude that, regardless of model
specification, geographic patterns of \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems \aprcomment{(Figure 3)}.

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (Supp.
Methods). The time of divergence of Western and Eastern African populations was
set to 60kya, just prior to the split of Eurasians and East Africans 50kya. Both
models predict relatively high subsequent gene flow between the two regions
($m=2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by ADMIXTURE
\aprcomment{cite} (Figure~\ref{fig:1}).

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{} % \includegraphics{figures/fig2.png}}
\caption{\textbf{}.
    A placeholder - best fit model(s).
}
\label{fig:2}
\end{center}
\end{figure}

@@

To account for population structure prior to 135kya, we allowed for two or more
“stem” populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) expansion of one or more of
the stem populations, with an instantaneous pulse (or “merger”) from the other
stem population. These models are meant to capture scenarios 2B and 2C (Figure
2). \aprcomment{update this last sentence}

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration (e.g. $LL \approx -102,600$ vs.
$-107,700$). With continuous migration, population structure extends back to
1.1--1.4Mya \aprcomment{(Table 1)}. Migration between the stems in these models
is low but non-negligible, with a fraction of migrant lineages each generation
estimated as $m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is
at least as large as migration inferred between the Ethiopian East Africans and
British populations over the past 50kya \aprcomment{(Table 1)}. This ongoing
(or, at least, periodic) gene flow qualitatively distinguishes these models
from previously proposed archaic admixture models \aprcomment{(Figure S1C)} as
the early branches remain closely related throughout (Figure 3).
\aprcomment{this last phrase may need to be updated, with regard to Fig
\ref{fig:supp-FST-panels}}

Likelihood-based parameter confidence intervals are relatively narrow within
each specified model
(Tables~\ref{tab:supp-single-origin-fits}--\ref{tab:supp-merger-with-stem-migration-fits}),
reflecting an informative statistical approach. As expected, model assumptions
have a larger impact on parameter estimates. To convey model uncertainty, we
highlight features of two models with the highest \aprcomment{not the
highest\ldots merger-without-stem-migration at -107,650}log likelihoods. These
differ primarily in whether early stems are assumed to give rise to modern
populations in multiple discrete merger events or in continuous migrations
\aprcomment{(Figure 2)}. Additional models are discussed in
\aprcomment{Supplement XXX}.

The “Multiple Merger” model has better log-likelihood than the “Continuous
Migration” model (-102,600 vs. -116,000, respectively). The two models differ
in the timing of divergence and the relative Ne of the two stems
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Under the continuous-migration model, the two stems give rise to contemporary
populations in western, southern and eastern Africa but with differing
proportions of ancestry derived from “Stem 1” and “Stem 2” (Figure 2). For
example, the Nama and Gumuz have lower levels of migration with Stem 2
($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), relative to the
Mende. The Mende continue to exchange migrants with Stem 2
($m=1.6\times10^{-4}$) long after their separation 60kya; this is a similar
level of gene flow as between the Mende and Eastern Africans (Table 1). 
\aprcomment{paragraphs above need work, and condensing}

\aprcomment{also below, too much information - we don't actually believe these
models}

In contrast, under the Multiple Merger model, we allowed for stem populations
to have large instantaneous contributions, in different proportions, to form
the different contemporary groups. However, the ghost populations do not neatly
line up one-to-one with southern, western and eastern populations. We observe a
deep bottleneck in Stem 1 down to $N_e=117$. This represents the lower bound
allowed in our optimization (i.e. 1\% of the $N_{ancestral}$) and the size of
this bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period
of exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
166$kya). These populations evolve independently until approximately 119kya
when Stem 1S and Stem 2 merge to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 merge to
form the ancestors of the Western Africans and Eastern Africans (and thus also
all individuals who later disperse during the Out of Africa event). The
ancestral Western / Eastern African group derives equal ancestry from the two
stems (50\% from Stem 1E and Stem 2, Table 1). Finally, the Mende receive a
large pulse of gene flow from Stem 2, replacing 18\% of their population 25kya.
The later Stem 2 contribution to the Western African Mende resulted in better
model fits (LL). This may indicate that an ancestral Stem 2 population
occupied Western or Central Africa, broadly speaking. The differing proportions
in the Nama and Eastern Africans may also indicate geographic separation of
Stem 1S in southern Africa and Stem 1E in eastern Africa.

\begin{table}
    \centering
    \begin{tabular}{l|l|l}
        \toprule
        Parameter & Value & Something else \\
        \midrule
        x & y & z \\
        \bottomrule
    \end{tabular}
    \caption{caption}
    \label{tab:migration-rates}
\end{table}

@@

Previous studies have found support for archaic admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. \aprcomment{Will need to check how strongly the Relate
paper claims that this supports archaic admixture vs simply deep structure. It
wasn't a model fit, in any case.} However, none of these studies considered a
weakly structured stem. We validated our Continuous Migration and Multiple
Merger models with additional independent approaches. We find that the two
models describe the cSFS accurately \aprcomment{(Figure SX)}, even though this
statistic was not used in the fit.  Both models outperform archaic models fit
directly to the cSFS (for example, compare with Figure~1 in
\citet{Durvasula2020-td}).

Similarly, we generated estimates of effective population size as inferred from
coalescence rates with Relate. Changes in Ne over time from prior studies
strongly support a bottleneck in Out of Africa populations, but African
populations also dip in effective size around 200kya, although this decrease is
more modest than the OOA event [Mallick 2016]. One possible explanation is that
ancestral population structure during the Middle Stone Age inflated Ne and
subsequent gene flow among ancestral Africans resulted in a lower Ne after
100-200kya [Chikhi?]. To our knowledge, explicit comparisons of this hypothesis
with SMC-style approaches have not been published. We simulated genomic data
under 4 models (Figure SX). All models, including the Single Origin model,
generate multi-modal curves indicating that the ancestral increase in Ne
between 100kya-1Mya is a general feature of the coalescent rate in humans.
Simulated Relate Ne curves under the CM and MM are distinguished by having 2
and 3 modes, respectively. The Continuous Migration model had the better
qualitative fit with the data, but the ancestral mode was shifted further back
in time relative to the data. All simulated models underestimated the level of
population growth during the Holocene, indicating additional free parameters
are necessary to refine these growth rates. \aprcomment{Discussion on this
paragraph moved to issues (issue \#1)}

Relative cross-coalescence rates (rCCR) have recently been used to estimate
divergence between a pair of populations, as measured by the rate of
coalescence between two groups divided by the mean within population
coalescence. Simulations of rCCR accuracy, however, focus on a ‘clean split’
between populations whereby groups diverge without subsequent gene flow.
Published estimates of the earliest human divergences with rCCR, which range
from 150kya-100kya \citep{Bergstrom2021-iw}, may be significantly biased when
compared to more complex models with gene flow as inferred here. We find that
midpoint estimates of rCCR are poor estimates for population divergence, often
underestimating divergence time by 50\% or greater (e.g. Mende vs. Gumuz ~15kya
compared to a true divergence 60kya, \aprcomment{Supp Figure SX}). We suggest
that rCCR analyses which do not fit multiple parameters including gene flow
should be interpreted with caution.

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human \bmhcomment{we use
this phrase a lot. human or hominin? Tim?} population structure, we sought to
capture model uncertainty by exploring multiple parameterizations of early
human population structure. The best-fit models presented here include
reticulation and low-level migration between early human populations rather
than archaic admixture. We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches (and therefore additional
parameters) during that period. \sgcomment{Some statement about hitting the
boundary} Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent, e.g., with a coalescence and
fragmentation model \citep{Scerri2019-xg}.  Eventual ancient DNA samples from
Africa could further distinguish the archaic admixture model from the weakly
structured stem model. \sgcomment{Or models involving more diversity within
Africa?}

By contrast, our inferred models paint a more consistent picture of the late
Middle Stone Age as a critical period of change, assuming that estimates from
the recombination clock accurately relate to geological chronologies
(Supplement). During the Middle Stone Age, the Multiple Merger indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with this
period of divergence, isolation and then a merger event (i.e. a “reticulation”)
out-performed models with bifurcating divergence and continuous gene flow.

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
modern groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
The second ~100kya between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later “Out of Africans”.  The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past ~120kya. These
subpopulations can be linked to modern-day groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Nama has an
XX probability of being in the ancestral ‘southern’ subpopulation 50kya versus
XX probability of being in the ‘eastern’ subpopulation (Figure XX). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table 1).

Evidence for archaic admixture in Eurasia has bolstered the plausibility of
archaic admixture having also occurred in Africa. For this reason, previous
work has focused on archaic admixture to explain patterns of polymorphism
inconsistent with a single origin model. Here, we have shown that weakly
structured stem models better capture these patterns. They also help explain an
ecological riddle posed by the archaic admixture model. Neanderthal populations
were separated from early \emph{Homo sapiens} by thousands of kilometers and
continental geographic barriers. By contrast, an archaic hominin population in
Africa would need to have stayed in relative reproductive isolation from the
human lineage over hundreds of thousands of years despite closer geographic
proximity and reproductive compatibility. The weakly structured stem model
resolves this ecological riddle by accommodating continuous or recurrent
contact between two or more groups present in Africa.

There is evidence for both deleterious and adaptive archaic-derived alleles in
modern genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-related
haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
\citet{Zhang2021-xx}).\sgcomment{Probably not the best reference} Under an
archaic African admixture model, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-derived haplotypes over a
short period \citep{Wall2019-ao}. By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly structured stem model in Africa.

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
assume that this correspondence also holds for early human ancestral
populations in Africa, the weakly structured stem model would predict that the
fossils deriving from each of the two stems should be morphologically quite
similar (supplementary text X)\aprcomment{Will we still include this?}, even if
they are from different parts of Africa, with differences across stems closer
to those across contemporary human populations than to human-Neanderthal
differences. Differences within stems would be comparable to differences within
recent human populations. Identifying the morphological signature of ancient
structure in the fossil record will require careful analysis of many broadly
similar remains. Our model further predicts that morphologically distinct
crania such as XX and YY are unlikely representatives of branches that
contributed appreciably to modern human ancestries.\sgcomment{Here I am being
somewhat forceful. We could use conditional to emphasize that these conclusion
are under the weakly structured stem model.}

\section*{Methods}

See Supplement, for now.

@@

We are grateful for the DNA contribution from each participant which enabled
this study; in particular we wish to highlight the generous participation of
the Richtersveld Nama community in South Africa and help from local research
assistants Willem DeKlerk and Hendrik Kaimann. Additional assistance and
community engagement was conducted by Justin Myrick, Chris Gignoux, Caitlen
Uren and Cedric Werely. We thank the African Genome Diversity Project for data
generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
also thank Sriram Sankararam, \dots for helpful discussion. This research was
supported by an NIH grant R35GM133531 (to BMH) and XXX. The content is solely
the responsibility of the authors and does not necessarily represent the
official views of the National Institutes of Health.
